On the spatial aggregation of ecosystem condition indicators

### 2024-11-20 manuscript/manuscript.qmd
First draft

---
title: On the spatial aggregation of ecosystem condition indicators
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Hanno Sandvik
    email: hanno.sandvik@nina.no
    orchid: https://orcid.org/0000-0002-5889-1606
    affiliations:
      - id: nina
        name: Norwegian Institute for Nature Research
        department: Department of Aquatic Biodiversity
        address: Pb 5685 Torgarden
        city: Trondheim
        postal-code: 7485
  - name: Bálint Czúcz
    email: balint.czucz@nina.no
    affiliations:
        - ref: nina
abstract: |
  Ecosystem condition assessments (ECA) and accounts use variables and indicators to describe key ecosystem characteristics reflecting the ecosystem condition and any potential deviations from a reference condition. These metrices are rutinely aggregation spatially to produce values representing the condition of a larger area, such as a country. However, we experience little awareness about the potential pitfals arising from aggregation bias of highly modified indicators. Here we outline some consequences regarding the order of the steps involed in normalising variables aggregating them in space. We show that aggregation bias in non-trivial and has the potential to undermine the credibility and precition in ECAs, and to confuse the communication of their results. We introduce a common terminology for aggregatio pathways, spesific to ecosystem condition indicators follwing SEEA EA standards, and make some recomendations about which aggregation pathway to use in different settings, and how to report these choices.

keywords: 
  - SEEA EA
  - ecosystem condition
  - ecosystem accounting
  - indicators
  - aggregation bias
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: ---
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: sentence
---

@@

In ecosystem accounting, fine-scale ecological information is routinely aggregated in space to produce single parameter estimates for entire regions.
This simplification is necessary to show major changes in the extent and condition of ecosystems at a scale that is relevant for decision makers.
The spatial aggregation of data can lead to biases arising from a number of sources, such as uneven sampling efforts, unequal sampling designs, or from the mathematical process of aggregating a variable that has a nonlinear relationship to a secondarily-derived metric (ref).
The first two sources of error are usually explicitly addressed and managed, but the last case of aggregation bias is often overlooked.
This error is most commonly encountered in ecosystem condition accounting when variables hold nonlinear relationships to the normalised indicators.
In the statistical standard for EAs and the recommended guidelines for ecosystem accounts (EAs) (ref), there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (a thematic aggregation), but in this paper we are discussing spatial aggregation only.

Variables in ecosystem condition terminology are metrics describing ecosystem characteristics.
They have a spatial component, tying them to unique Ecosystem Assets.
Variables can be spatially aggregated across ecosystem assets (or across basal units), for example by taking the sum or an area-weighted mean, to produce variable estimates for larger regions, such as the entire Ecosystem Accounting Area (EAA).
As a next step, variables can be normalised and turned into normative indicators for ecosystem condition.
The mathematical part of this normalisation contains two, sometimes three, steps [@fig-term].
These steps can be performed simultaneously, but for clarity we discuss them as separate steps.
First the variable is scaled, using at least two reference levels, defining variable values to be coded as zero (X0) or one (X100) on the indicator scale.
Secondly, the scaled variables are truncated to produce a bound indicator scale between 0 and 1.
Sometimes a third step of transformation is required to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
This is commonly done by anchoring specific variable values to predefined class boundaries (ref SELINA paper), but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
When a variable is normalised in this way, its interpretation changes from descriptive to normative.
The ecological significance of the indicator might also differ from that of the variable since additional ecological knowledge can be introduced via the reference values.
The indicator is a function of the variable and the reference values, and this relationship is almost always nonlinear (e.g. due to truncation which is ubiquitous practice).

Similar to variables, the reference values used for normalising the variables also have a spatial assignment, but it can be different from that of the variable.
For example, variable values may exist for unique 10x10m grid cells, but the reference values may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire EAA (e.g. a natural zero).
A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition variables or indicators, is the choice about the order of the different steps in the normalisation and aggregation process, i.e. the aggregation pathway (Figure 2).

```{r fig-term}
#| fig-cap: |
#|   Examples of how important terms are used in this manuscript. The variable (1), in raw biophysical units, has a minimum and a maximum value (V), as well as points anchoring it to the indicator scale (X). Scaled variables (2) are scaled based on X0 and X100 only. Note that this is a different use of the term then the more common use that implies dividing a vector by its standard deviation. Truncation (3) implies assigning the value of 0 to values below X0, and the value 1 to value above X100. Transformation refers to the non-linear transformation of values within the 0-1 range, as in this piecewise-linear example by mapping X60 to 0.6 on the indicator scale. Transformations without anchoring points are also included in this term, such as exponential or sigmoid transformation. Although both truncation and scaling can be seen as types of transformations, we exclude these methods from the definition here. Examples 2, 3, and 4 are normative (hence the gradient colour scale from red to green) because they have declared that one end of the variable scale represents a good state and the other a poor state. All variables that are treated so that they conform to the definition of indicators in the SEEA EA are said to be normalised. For example, if a truncated variable is assumed to have a linear relationship with the indicator scale, then this variable has been normalised even though it has not been transformed. The term rescale is a synonym to normalise, but we will only use the latter from here on."
#| echo: false
knitr::include_graphics("../img/terminologi.png")
```

```{r fig-pathways}
#| fig-cap: |
#|   "Flowchart showing pathways for the spatial aggregation of variables in ecosystem condition accounting. Normalisation refers to the steps that make a variable conform to the definition of an indicator in SEEA EA, which may or may not include the transformation step. A truncated and/or transformed variable will have a different interpretation than the original variable. The aggregation bias depends on whether variables are truncated and/or transformed variables before or after spatial aggregation. Variables are never normalised (pathway 4), and hence the aggregation bias for variables will be different than for indicators. An important factor influencing the the choice whether to normalise early (as in pathway 1), or later, is the spatial validity of the reference levels, and sometimes raw variable values (e.g. billberry coverage in vegetation quadrats) are spatially aggregated to the scale that was considered when setting the reference levels (e.g. average regional billberry coverage)."
#| out-width: 400
#| echo: false
knitr::include_graphics("../img/aggregationPathways.jpg")
```

```{r fig-agg1}
#| echo: false
#| fig-cap: |
#|   "Example showing the difference in indicator values (y-axis) obtained from taking the mean of two variable values (V1 and V1) that are on the original scale (i hat) or on the normalised indicator scale (i bar). Due to truncation at x100, the latter results in a comparatively lower indicator value. The solid coloured line represents the normalisation function f(V). V = variable value, i = indicator value. Modified from Rastsetter (1991)."
#| out-width: 400

@@

```{r fig-agg2}
#| echo: false
#| fig-cap: |
#|   "Same figure as above, but showing an example where the normalisation function includes a transformation step (see Figure 1 for definition) that anchors X60 to 0.6 on the indicator scale. "
#| out-width: 400

@@

## Why do we normalise? {#sec-norm}

Normalisation of ecosystem condition indicators, as defined in Figure 1, serves multiple purposes.
Firstly, it gives a normative interpretation of a variable, defining a good and a bad state and a directionality to say when something is getting better or worse over time.
This normative characteristic is an essential part of assessing ecosystem condition.
Secondly, normalisation sets a limit to how much a high indicator value in one place can compensate for a low value somewhere else.
This is because the truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviation from the reference levels, and positive deviations (which could compensate for the negative ones) are ignored.
This is sometimes what we want, but not always something we are aware that we are doing.
One reason to want to aggregate the negative deviations only, is because the reference levels are set (or should be set) so that values above these limits do not represent any further increase (or decrease) in ecosystem condition.
Therefore, this way of aggregation summarises the presumed ecosystem condition (which is generally what we want), and not the variable itself.
In other words, we normalise in order to facilitate the spatial (i.e. horizontal) aggregation of our ecosystem condition estimates (i.e. the indicator values), and not merely the variable values.
Thirdly, we normalise in order to standardise the indicator on the same scale so that we can perform thematic (i.e. vertical) aggregation.
This is commonly referred to as the reason for normalising variables, but as we have shown, it is but one of three main reasons, and also perhaps the least confusing part for many.

## Main points to consider when deciding on aggregation pathway

### Ease of interpretation {.unnumbered}

The indicator should make intuitive sense.
This is helped by using the same aggregation pathway for all indicators.
Indicators using spatially aggregated population estimates are easily misinterpreted as representing population trends.

### The ‘real’ spatial resolution of reference values {.unnumbered}

To achieve an early normalisation of variables, we need reference values that make ecological sense at the same spatial resolution as the variable.
Spatial disaggregation or interpolation of reference values needs to be well justified.

### Scalability and comparability between assessment {.unnumbered}

Separate assessments should give identical indicator values for the same areas, given the same underlying data.
When this is not true it raises concerns about reproducibility and sacrifices the credibility of both assessments.

### Making clear the differences in the interpretation of variables and indicators {.unnumbered}

Often indicators and variables are both included side-by-side in ecosystem condition assessments/accounts.
This is also recommended practice following the SEEA EA. However, if the indicators that are presented are normalised at an earlier stage, and subsequently spatially aggregated, same as the variable, then the interpretation of the indicators is not simply a normalised version of the variable, but rather a whole new beast.
For example, these indicators typically now reflect the average ecosystem condition, where areas in very good or very poor condition have limited ability to compensate for opposite extreme values elsewhere.
The spatially aggregated variable on the other hand, may reflect something like a sum of individuals, or some other aspect where values above and below the average value are able to compensate for each other.

## Examples

In this section we use some examples from our own experience to illustrate some of the issues arising from chosing different aggregation pathways,

### Billberry example

@@

@@

### Phytoplankton trophic index (PTI)

@@

>>>>

---
title: On the spatial aggregation of ecosystem condition indicators
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Hanno Sandvik
    email: hanno.sandvik@nina.no
    orchid: https://orcid.org/0000-0002-5889-1606
    affiliations:
      - id: nina
  - name: Bálint Czúcz
    email: balint.czucz@nina.no
    affiliations:
        - ref: nina
abstract: |
  Ecosystem condition assessments (ECA) and accounts use variables and indicators to describe key ecosystem characteristics reflecting the ecosystem condition and any potential deviations from a reference condition. These metrices are rutinely aggregation spatially to produce values representing the condition of a larger area, such as a country. However, we experience little awareness about the potential pitfals arising from aggregation bias of highly modified indicators. Here we outline some consequences regarding the order of the steps involed in normalising variables and aggregating them in space. We show that the choice of  aggregation pathways in non-trivial and has the potential to undermine the credibility and precition in ECAs, and to confuse the communication of their results. We introduce a common terminology for aggregation pathways, spesific to ecosystem condition indicators following SEEA EA standards, and make some recomendations about which pathway to use in different settings, and how to report these choices.

keywords: 
  - SEEA EA
  - ecosystem condition
  - ecosystem accounting
  - indicators
  - aggregation bias
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: ---
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
execute:
  cache: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: sentence
---

@@

In ecosystem accounting (EA), fine-scale ecological information is routinely aggregated in space to produce single parameter estimates for entire regions.
This simplification is necessary to show major changes in the extent and condition of ecosystems at a scale that is relevant for decision makers.
The spatial aggregation of data can lead to biases arising from a number of sources, such as uneven sampling efforts, unequal sampling designs.
In ecosystem condition accounting (ECA) specifically, the aggregation of both variables and their normative indicators is common practice and raises the issue of aggregation bias when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
The first two sources of error are usually explicitly addressed and managed, but the last case of aggregation bias is often overlooked.
In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2021; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (a thematic aggregation), but in this paper we are discussing spatial aggregation only.

Variables in ecosystem condition terminology are metrics describing ecosystem characteristics [@unitednations2021].
They have a spatial component, tying them to concrete areas, which can be of varying size, from single pixels to entire regions.
Variables can be spatially aggregated, for example by taking the sum or an area-weighted mean, to produce variable estimates for larger regions, such as the entire Ecosystem Accounting Area (EAA).
Variables can also be normalised and turned into normative indicators for ecosystem condition.
The mathematical part of this normalisation contains two, sometimes three, steps [@fig-term].
These steps can be performed simultaneously, but for clarity we discuss them as separate steps.
The order of these steps can also change.
Variable can be scaled using at least two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
Variables can be truncated to produce a bound indicator scale between 0 and 1.
Sometimes variables are transformed to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
When a variable is normalised in this way, its interpretation changes from descriptive to normative.
The ecological significance of the indicator might also differ from that of the variable since additional ecological knowledge can be introduced via the reference leveles.
The indicator is a function of the variable and the reference values, and this relationship is almost always nonlinear (e.g. due to truncation which is ubiquitous practice).

Similar to variables, the reference values used for normalising the variables also have a spatial assignment, but it can be different from that of the variable.
For example, variable values may exist for unique 10×10m grid cells, but the reference values may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire EAA (e.g. a natural zero).
A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition indicators (@fig-agg1; @fig-agg2), is the choice about the order of the different steps in the normalisation and aggregation process, i.e. the aggregation pathway (@fig-pathways).

```{r fig-term}
#| fig-cap: |
#|   Examples of how important terms are used in this manuscript. Note that the meaning of the terms may differ from how they are used in other fields, such as mathematics. The variable (1), in raw biophysical units, has a minimum and a maximum value (V), as well as points anchoring it to the indicator scale (X). Scaled variables (2) are scaled based on X~0~ and X~100~ only. Truncation (3) implies assigning the value of 0 to values below X~0~, and the value 1 to value above X~100~. Transformation refers to the non-linear transformation of values within the 0-1 range, as in this piecewise-linear example by mapping X~60~ to 0.6 on the indicator scale. Transformations without anchoring points are also included in this term, such as exponential or sigmoid transformation. Although both truncation and scaling can be seen as types of transformations, we exclude these methods from the definition here. Examples 2, 3, and 4 are normative (hence the gradient colour scale from red to green) because we assume they have declared that one end of the variable scale represents a good state and the other a poor state. All variables that are treated so that they conform to the definition of indicators in the SEEA EA are said to be normalised. For example, if a truncated variable is assumed to have a linear relationship with the indicator scale, then this variable has been normalised even though it has not been transformed. The term rescale is a synonym to normalise, but we will only use the latter from here on.
#| echo: false
knitr::include_graphics("../img/terminologi.png")
```

# Why do we normalise? {#sec-norm}

Normalisation of ecosystem condition indicators, as defined in @fig-term, serves multiple purposes.
Firstly, it gives a normative interpretation of a variable, defining a good and a bad state and a directionality to say when something is getting better or worse over time [@czucz_selection_2021].
<!--# This normative characteristic is an essential part of assessing ecosystem condition. --> Secondly, normalisation sets a limit to how much a high indicator value in one place can compensate for a low value somewhere else, and vice versa.
This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviation from the reference levels, and positive deviations (which could compensate for the negative ones) are ignored.
Transformation can sometimes also have this effect, e.g. sigmoid transformations.
<!--# This is often what we want (but not always), but not always something we are aware that we are doing. -->One reason to want to aggregate the negative deviations only, is because the reference levels are set (or should be set) so that values above these limits do not represent any further increase (or decrease) in ecosystem condition.
(An exception if the WFD; see @sec-pti).
Therefore, this way of aggregation summarises the estimated ecosystem condition (which is generally what we want), and not the variable itself.
In other words, we normalise in order to facilitate the spatial (i.e. horizontal) aggregation of our ecosystem condition estimates (i.e. the indicator values), and not merely the variable values.
Thirdly, we normalise in order to standardise the indicator on the same scale so that we can perform thematic (i.e. vertical) aggregation.
This is commonly referred to as the reason for normalising variables, but as we have shown, it is but one of three main reasons, and also perhaps the least confusing part for many.

```{r fig-agg1}
#| echo: false
#| fig-cap: |
#|   Example showing the difference in indicator values (y-axis) obtained from taking the mean of two variable values (V1 and V1) that are on the original scale (i hat) or on the normalised indicator scale (i bar). Due to truncation at x100, the latter results in a comparatively lower indicator value. The solid coloured line represents the normalisation function f(V). V = variable value, i = indicator value. Modified from Rastsetter (1991).
#| out-width: 400

@@

```{r fig-agg2}
#| echo: false
#| fig-cap: |
#|   Same figure as above, but showing an example where the normalisation function includes a transformation step (see Figure 1 for definition) that anchors X60 to 0.6 on the indicator scale.
#| out-width: 400

@@

# Aggregation pathways

We introduce the term aggregation pathway to describe the order of the steps used to go from a spatially explicit variable to a spatially aggregated metric, usually an indicator.
In @fig-pathways we show some aggregation pathways that we have come across, and some that we we see as potential new pathways.

Pathway 1 involves early normalisation suing the perhaps most common order of the three steps scaling, truncating and transforming.
The three steps may be done simultaneously as well.
Then the indicator is spatially aggregated.
Pathway 2 involves aggregating the variable before normalising.
Pathway 3 is similar, but here there are two aggregation steps: one before and one after normalisation.
This can for example be the case when variables are aggregated to the scale corresponding to the reference levels before they can be normalised.
Pathway 4 illustrated the aggregation of a variable, with no normalisation.
This is the pathway commonly used for variable accounts in the SEE EA. Pathway 5 is a common pathway in the WFD, which does not include spatial aggregation of indicator values.
Instead one report the number of water bodies in each condition class (spatial statistics).
Pathway 6 is a suggested pathway for handling WFD indicator in ECAs (see @sec-pti).

Pathways 7-8 describe the spatial aggregation of multiple indicators, as they all start with the normalisation step.
These are not the main topic of this paper, but are included here for completeness.
In pathway 7, indicators are thematically aggregated to an index, which is then again aggregated spatially.
In pathway 8, spatial aggregation is only done on individual indicators.
The two pathways differ in how they handle more complicated schemes for weighted averages, with pathway 8 being more flexible, but it may interfere with the principle of commutativity [@unitednations2021].

Pathway 9 describe how a single indicators is aggregated hierarchically from region to region, whereas pathway 10 shows a case were the aggregation process always starts from scratch for each hierarchical spatial level.
As for pathways 7 and 8, pathway 9 and 10 differ in how they propagate the indicator weight (typically area weight) between regions.

The choice about which pathway to use is not trivial, and in this paper we want to highlight some of the issues that could arise from having an *ad hoc* approach to these considerations.
Here we go through selected examples to illustrate some common pitfalls.

<!--# what we see as four general points that developers of ecosystem condition indicators and assessments should reflect upon carefully when deciding on an aggregation pathway for their indicator or their assessment. -->

## Wolves example {#sec-wolves}

*Variable*: Number of wolves en each carnivore management area (CMA).

*Reference:* X~0~ = is zero wolves present.
X~100~= Number of wolves equal to what experts think the ecosystem would support under the reference condition.
The reference levels are unique to each (CMA).

This example is from Norway, where wolves are extinct in most of the country.
Therefore, most carnivore management areas will get an indicator value of 0.
Wolves exists i designated wolf zones in south-east Norway, although in small numbers.
In the Norwegian Natur Index [@jakobsson2020], the variable follows pathway 1 and is normalised at the CMA level.
When aggregating the indicator to a national value, it is common for the general public to interpret this value as the status for wolves in Norway, and to connect this value to Norway's international obligations to maintain wolf numbers.
But note that even an unlimited number of wolves in the wolf zone would not to bring the national indicator value up to a level indicator good condition.
This is because the normalisation process truncates the variable, and does not allow overshooting values to compensate for lower values in the rest of the country.

\*Conclusion\*: Indicators should make intuitive sense to the readers.
However, there are sometimes good reasons for making things complicated.
In this example, The Norwegian Nature Index uses the same pathway for all indicators, so this common missunderstanding about what indicator values represenets can be exaplin just one time, and not for each indicator spesifically.
It can also be a good idea to sometime go back to the variable, and see if it can be defined differently to avoid these kinds of confusions.
Wolf numbers could in this example be converted to wolf density.

## Billberry example {#sec-bill}

@@

The wolves example also highlight the importance of making clear the differences in the interpretation of variables and indicators.
Often indicators and variables are both included side-by-side in ecosystem condition assessments/accounts.
This is also recommended practice following the SEEA EA. However, if the indicators that are presented are normalised before any spatial aggregation, then the interpretation of the indicators is not simply as a normalised version of the variable.
Indicators typically now reflect the average ecosystem condition, where areas in very good or very poor condition have limited ability to compensate for opposite extreme values elsewhere.
The spatially aggregated variable on the other hand, may reflect something like a sum of individuals, or some other aspect where values above and below the average value are able to compensate for each other.

@@

Separate assessments should give identical indicator values for the same areas, given the same underlying data.
When this is not true it raises concerns about reproducibility and sacrifices the credibility of both assessments.
This was also the case in the billberry example, but here we used another example to illustrate the point further.

### Phytoplankton trophic index (PTI) {#sec-pti}

@@

```{r fig-pathways}
#| fig-cap: |
#|   Flowchart showing pathways for the spatial aggregation of variables in ecosystem condition accounting. Normalisation refers to the steps that make a variable conform to the definition of an indicator in SEEA EA, which may or may not include the transformation step. A truncated and/or transformed variable will have a different interpretation than the original variable. The aggregation bias depends on whether variables are truncated and/or transformed variables before or after spatial aggregation. Variables are never normalised (pathway 4), and hence the aggregation bias for variables will be different than for indicators. An important factor influencing the the choice whether to normalise early (as in pathway 1), or later, is the spatial validity of the reference levels, and sometimes raw variable values (e.g. billberry coverage in vegetation quadrats) are spatially aggregated to the scale that was considered when setting the reference levels (e.g. average regional billberry coverage).
#| out-width: 400
#| echo: false
knitr::include_graphics("../img/aggregationPathways.jpg")
```

# Recommendation

Based on the example above, we suggest some recommendations for developers of ecosystem condition indicators and assessment for trying to avoid some of the pitfalls from having a too casual approach to the choice of aggregation pathways.

### 1. Always report the aggregation pathway, using standardised terminology {.unnumbered}

In indicator documentation or ECA repports, authors should make an effort to make a precise description of the steps in the aggregation process.
To do this we suggest using the terms as described in @fig-term.
The description can be made even easier by referencing specific aggregation pathways in @fig-pathways by number.
Besides making the indicator workflow itself more transparrent and reproduciable, ee believe this recommendation would help raise the general awareness about aggregation bias in EAs, both among readers and researchers.
It will make the interpretation of assessment less prone to misunderstanding.
It will also make troubleshooting easier, as i the billberry example where researchers had to spend several days looking though old code to identify the reason why two assessment produces dissimilar indicator values from the same underlying data.

### 2. Always normalise variables at the scale where the reference levels are relevant {.unnumbered}

Normalising variables early in the aggregation pathway has several benefits, as we have seen.
But same as variables and indicators, reference levels also have a spatial resolution, and they should only be used to normalise variables when the variable is at a scale which is relevant to the way the reference levels are defined.
This was exeplified with the billberry example (@sec-bill).

### 3. Use the same aggregation pathway for all indicators in the same assessment {.unnumbered}

As we saw in the wolves example (@sec-wolves) this eases interpretation and reduces the change that the meaning of indicators are misunderstood.
It may be premature, or not even possible, to prescribe an aggregation pathway to be used for all ECA, but internal consistency within assessment should be possible.

### 4. Use unique indicator IDs on indicators that are similar, but use different pathways {.unnumbered}

Indicators (and variables) often exist in multiple version, varying slightly in the raw data or in the methods used to produce the data or metrics.
Yet different versions are often refered to by the same common name.
This causes confusion about which indicator version is being used, and thus making it difficult to make out which aggregation pathway that has been used.
We recommend making use of stable indicator IDs, unique for each version of the indicators sharing the same common name.
This has for example been implemented in [ecRxiv](https://github.com/NINAnor/ecRxiv), a publishing platform for ecosystem condition indicators.
